Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-08-01 paper.tex
Improve text in the oscillation section

\subsection{Neutrino Oscillations at High Neutrino Densities}
\label{ssec:V_nunu}
The Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}) is one limiting
form of the quantum kinetic equations \cite{vlas2014-neutrino_qkes},
the limit where collisional mean free paths are much shorter
than oscillation lengths (i.e.\ $\varepsilon/\mathscr{K} \ll |H|^{-1}$
with $H$ the Hamiltonian describing coherent forward scattering interactions).
In this limit the neutrino density matrix,
\begin{equation}
  \label{eqn:density_rho_0}
  \rho_0(\varepsilon,\ell_\mu) =
  \sum\limits_{\alpha}f^{\nu_\alpha}(\varepsilon,\ell_\mu)
  \ket{\nu_\alpha} \bra{\nu_\alpha},
\end{equation}
with $\ket{\nu_\alpha}$ the neutrino flavor eigenstates,
remains in a pure state, or diagonal in the flavor basis.

But at the opposite limit, in the free-streaming regime,
coherence effects become important,
and the quantum kinetic equations take a Schr\"{o}dinger-like form:
\begin{equation}
  \label{eqn:schroedinger}
  i \frac{d}{ds} S = H S,
\end{equation}
where $ds$ is an interval of proper length, and
\todo{mention $\bar{\nu}$ eqn}
$S$ is the flavor evolution matrix describing a mixed state
\begin{equation}
  \label{eq:density_evolution}
  \rho = S \, \rho_0 \, S^\dagger.
\end{equation}
In this limit, we may decompose the Hamiltonian into vacuum,
matter, and neutrino contributions:
\begin{equation}
  \label{eqn:decompose_H}
  H = H_{\rm V} + H_e + H_{\nu\nu}.
\end{equation}
Explicit formulas for these terms may be found in the oscillation literature,
e.g.\ \cite{duan2009-review},
but for simplicity we give their respective scales:
\begin{align}
  \label{eqn:scale_H_V}
  |H_{\rm V}| &\sim \frac{\Delta m^2}{\varepsilon}, \\
  \label{eqn:scale_H_e}
  |H_e|       &\sim G_{\rm F}\left|n_{e^-}-n_{e^+}\right|, \\
  \label{eqn:scale_H_nunu}
  |H_{\nu\nu}(\ell_\mu)|&\sim G_{\rm F} \left|
  (G_{\nu_e}-G_{\bar{\nu}_e}) - \ell_\mu(K_{\nu_e}^\mu-K_{\bar{\nu}_e}^\mu)
  \right|,
\end{align}
where $\Delta m$ is the mass difference between neutrino mass eigenstates,
$G_{\rm F}$ is the Fermi coupling constant,
$n_{e^-}$ and $n_{e^+}$ are the electron and positron number densities,
and $\ell_\mu$ is the propagation direction of the
coherent beam of neutrinos in question.

@@

However, we can solve a similar, but tractable problem along a single ray.
Assume all the neutrino rays intersecting an event along a given test ray
have undergone the same flavor evolution history as that of the test ray,
i.e.\ $S$ is the same for all rays sharing that event.
This is the so-called single-angle approximation.
\todo{cite someone}
\todo{say something about its validity}

Solving for the flavor evolution history along this test ray
requires knowledge of the unoscillated neutrino contribution to the Hamiltonian.
This is a function of the unoscillation neutrino self-interaction potentials,
which for the $\alpha$-th flavor is:
\begin{equation}
  \label{eqn:V_nu_alpha}
  V_{\nu_\alpha,0}(\varepsilon,\ell_\mu) =
  \frac{\sqrt{2}G_{\rm F}}{(2\pi)^3} \varepsilon^2
  \oint d\Omega'(1-\varpi')f^{\nu_\alpha}(\varepsilon,\ell'_\gamma),
\end{equation}
with the cosine of the angle between the test ray and ambient rays
given by $\varpi'=\psi^{\mu\gamma} \ell_\mu \ell'_\gamma$.

If conditions are right, a resonant flavor transition
introducing significant mixing may occur very near
the point where neutrinos begin free-streaming.
The matter-neutrino resonance \cite{malk2012-mnr_1,malk2015-mnr_2,malk2016-mnr_3},
may occur along a given ray wherever the matter and neutrino self-interaction
potentials cancel, or $V_e \sim V_{\nu\nu,0}$, with
\begin{align}
  \label{eqn:V_e}
  V_e &= \sqrt{2} G_{\rm F} \frac{\varrho}{m_N} Y_e, \\
  \label{eqn:V_nunu0}
  V_{\nu\nu} &= \int d\varepsilon
  \big(V_{\nu_e,0}(\varepsilon)-V_{\bar{\nu}_e,0}(\varepsilon)\big),
\end{align}
with $\varrho$ the rest density and $m_{N}$ the nucleon mass.

We calculate the self-interaction potential at $N_s=5$ points between
$r\in[30,135]$~km,
by sampling distribution functions over a grid with extents
$N_\varepsilon=30$, $N_A=30$, $N_B=10$,
over all angles and an energy range of $\varepsilon\in(0,100)$~MeV.
\todo{improve resolution and update}
We interpolate the logarithm of the self-interaction potentials
in path length $\log s$ along the ray,
by fitting a 3rd order spline with continuous derivative,
across the $N_s$ observation points.
The matter potential along this test ray is from an analytic wind
model qualitatively consistent with the densities in the
simulated volume (i.e.\ inside $r\sim70$~km).

@@

Evolving the flavor content of this system using the formalism described in
\cite{zhu2016-mnr_nsns_remnant}
we see that the matter-neutrino resonance occurs in both,
but with the very different behaviors of symmetric and asymmetric
resonances respectively for the \emph{noscat} and \emph{scat} cases,
as seen in Figs.~\ref{fig:survival-noscat} and \ref{fig:survival-scat}.
These two classes of matter-neutrino resonance
are described e.g.\ in \cite{malk2016-mnr_3,vaan2016-uncovering_mnr}.
The survival probabilities shown in these figures are the terms of
the flavor evolution matrix $S$ that describe how much of the original flux
of a given flavor survives a flavor transition.
Where $P_{\nu_e\rightarrow\nu_e}$ vanishes, for example in
Fig.~\ref{fig:survival-noscat}, all the original $\nu_e$ content of the beam
is oscillated into $\nu_\mu$ and $\nu_\tau$; but the $\nu_e$ density doesn't
vanish altogether,
since some $\nu_\mu$ and $\nu_\tau$ oscillates into $\nu_e$.
This dramatic difference in survival probabilities between the \emph{noscat}
and \emph{scat} cases along this test ray emphasizes the importance of
scattering physics in the problem of neutrino oscillations in
neutron star mergers.

>>>>

\subsection{Neutrino Oscillations at High Neutrino Densities}
\label{ssec:V_nunu}
The Boltzmann Equation (Eqn.~\ref{eqn:boltzmann}) is one limiting
form of the quantum kinetic equations \cite{vlas2014-neutrino_qkes},
the limit where collisional mean free paths are much shorter
than oscillation lengths (i.e.\ $\varepsilon/\mathscr{K} \ll |H|^{-1}$
with $H$ the Hamiltonian describing coherent forward scattering interactions).
In this limit the neutrino density matrix takes the form
\begin{equation}
  \label{eqn:density_rho_0}
  \rho_0(\varepsilon,\ell_\mu) =
  \sum\limits_{\alpha}f^{\nu_\alpha}(\varepsilon,\ell_\mu)
  \ket{\nu_\alpha} \bra{\nu_\alpha},
\end{equation}
with $\ket{\nu_\alpha}$ the neutrino flavor eigenstates.
The neutrino state remains pure, or diagonal in the flavor basis.

But at the opposite limit, in the free-streaming regime,
coherence effects become important,
and the quantum kinetic equations take a Schr\"{o}dinger-like form:
\begin{equation}
  \label{eqn:schroedinger}
  i \frac{d}{ds} S = H S,
\end{equation}
where $ds$ is an interval of proper length, and
\todo{mention $\bar{\nu}$ eqn}
$S$ is the flavor evolution matrix describing a mixed state
\begin{equation}
  \label{eq:density_evolution}
  \rho = S \, \rho_0 \, S^\dagger.
\end{equation}
In this limit, we may decompose the Hamiltonian into vacuum,
matter, and neutrino contributions:
\begin{equation}
  \label{eqn:decompose_H}
  H = H_{\rm V} + H_e + H_{\nu\nu}.
\end{equation}
Explicit formulas for these terms may be found in the oscillation literature,
e.g.\ \cite{duan2009-review},
but for clarity we give their order-of-magnitude scales here:
\begin{align}
  \label{eqn:scale_H_V}
  |H_{\rm V}| &\sim \frac{\Delta m^2}{\varepsilon}, \\
  \label{eqn:scale_H_e}
  |H_e|       &\sim G_{\rm F}\left|n_{e^-}-n_{e^+}\right|, \\
  \label{eqn:scale_H_nunu}
  |H_{\nu\nu}|&\sim G_{\rm F} \left|G_{\nu_e}-G_{\bar{\nu}_e}\right|,
\end{align}
where $\Delta m$ is the mass difference between neutrino mass eigenstates,
$G_{\rm F}$ is the Fermi coupling constant,
$n_{e^-}$ and $n_{e^+}$ are the electron and positron number densities,
and $G_{\nu_e}$ and $G_{\bar{\nu}_e}$ the electron neutrino and antineutrino
number densities defined in Eqn.~\ref{eqn:G}.
For the scale of $H_{\nu\nu}$ we have only included the isotropic
components of the neutrino fields.

@@

However, we can solve a similar but tractable problem along a single ray.
Assume all the neutrino rays intersecting an event along a given test ray
have undergone the same flavor evolution history as that of the test ray,
i.e.\ $S$ is the same for all rays sharing that event.
This is the so-called single-angle approximation.
\todo{cite someone}
\todo{say something about its validity}
Though widely used in the supernova oscillation literature
and shown to be qualitatively faithful in those environments,
\todo{cite vlasenko?}
a recent study has shown that the single-angle approximation
may mask certain instabilities in the flavor evolution
\cite{wu2017-fast_neutrino_conversions}.

Within the single-angle approximation,
solving for the flavor evolution history along a given test ray
requires knowledge of the unoscillated neutrino contribution to the Hamiltonian
along that trajectory.
This is a function of the unoscillated neutrino self-interaction potential,
which for the $\alpha$-th flavor is:
\begin{equation}
  \label{eqn:V_nu_alpha}
  V_{\nu_\alpha,0}(\varepsilon,\ell_\mu) =
  \frac{\sqrt{2}G_{\rm F}}{(2\pi)^3} \varepsilon^2
  \oint d\Omega'(1-\varpi')f^{\nu_\alpha}(\varepsilon,\ell'_\gamma),
\end{equation}
with the test ray propagating in direction $\ell_\mu$,
ambient rays propagating in directions $\ell'_\gamma$,
and the cosine of the angle between these given by
$\varpi'=\psi^{\mu\gamma} \ell_\mu \ell'_\gamma$.
As in the moment equations (Eqns.~\ref{eqn:J}-\ref{eqn:Ka}),
the integral is taken over all directions $\ell'_\gamma$.

If conditions are right, a resonant flavor transition
introducing significant mixing may occur very near
the point where neutrinos begin free-streaming.
The matter-neutrino resonance
\cite{malk2012-mnr_1,malk2015-mnr_2,malk2016-mnr_3},
can occur where the matter potential and
neutrino self-interaction potentials cancel,
or where $V_e - V_{\nu\nu,0}=0$, with
\begin{align}
  \label{eqn:V_e}
  V_e &= \sqrt{2} G_{\rm F} \frac{\varrho}{m_N} Y_e, \\
  \label{eqn:V_nunu0}
  V_{\nu\nu} &= \int d\varepsilon
  \big(V_{\nu_e,0}(\varepsilon)-V_{\bar{\nu}_e,0}(\varepsilon)\big),
\end{align}
with $\varrho$ the rest density and $m_{N}$ the nucleon mass.
The matter potential is always positive;
and outside the accretion disks formed in neutron star mergers,
in which the disrupted neutron-star matter is rapidly releptonizing,
the neutrino self-interaction potential is large and negative.

We examine this effect in the post-merger configuration already analyzed
in Sec.~\ref{ssec:test_disk_comparison}.
We calculate the self-interaction potential at $N_s=5$ points between
$r\in[30,135]$~km,
by sampling distribution functions over a grid with extents
$N_\varepsilon=30$, $N_A=30$, $N_B=10$,
over all angles and an energy range of $\varepsilon\in(0,100)$~MeV.
\todo{improve resolution and update}
We interpolate the logarithm of the self-interaction potentials
(Eqn.~\ref{eqn:V_nu_alpha})
in path length $\log s$ along the ray,
by fitting a 3rd order spline with continuous derivative,
across the $N_s$ observation points.
The matter potential along this test ray is from an analytic wind
model qualitatively consistent with the densities in the
simulated volume (i.e.\ inside $r\sim70$~km).

@@

We also solve for the flavor evolution of this system,
integrating $S$ according to Eqn.~\ref{eqn:schroedinger},
as described in \cite{zhu2016-mnr_nsns_remnant}.
In Figs.~\ref{fig:survival-noscat} and \ref{fig:survival-scat}
we show the survival probabilities for $\nu_e$ and $\bar{\nu}_e$
for the two treatments.
The survival probabilities are the terms of
the flavor evolution matrix $S$ that describe how much of the original flux
of a given flavor survives a transition.
\todo{define explicitly}
Where $P_{\nu_e\rightarrow\nu_e}$ vanishes, for example in
Fig.~\ref{fig:survival-noscat}, all the original $\nu_e$ content of the beam
is oscillated into $\nu_\mu$ and $\nu_\tau$; but the $\nu_e$ density doesn't
vanish altogether,
since some $\nu_\mu$ and $\nu_\tau$ oscillates into $\nu_e$.

We see that the matter-neutrino resonance occurs in both treatments,
but with the very different behaviors of symmetric and asymmetric
resonances respectively for the \emph{noscat} and \emph{scat} cases,
as seen in Figs.~\ref{fig:survival-noscat} and \ref{fig:survival-scat}.
These two classes of matter-neutrino resonance
are described for example in \cite{malk2016-mnr_3,vaan2016-uncovering_mnr}.
This dramatic difference in survival probabilities between the \emph{noscat}
and \emph{scat} cases along this test ray emphasizes the importance of
scattering physics in the problem of neutrino oscillations in
neutron star mergers.
